---
title: "Project Adjusted Means"
author: "Alex Salem"
date: "December 2, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading libraries
```{r, message = FALSE}
library(tidyverse)
library(skimr)
library(readxl) # to read in the xlsx file
library(naniar) #for replacing 6 with NA
# install packages needed for ANCOVA analysis as per the "Discovering Statistics Using R" book by Field, Miles and Field
#install.packages("car"); install.packages("akima"); install.packages("compute.es"); install.packages ("effects");install.packages("ggplot2");install.packages("multcomp");install.packages("pastecs"); install.packages("WRS", repos="http://R-Forge.R-project.org")
# load ANCOVA relate libraries after installing them
library(car); # for Levene’s test, Type III sums of squares
library(compute.es); # for effect sizes
library(effects); # for adjusted means
library(ggplot2); # for graphs
library(multcomp); # for post hoc tests
library(pastecs); # for descriptive statistics
library(WRS) # for robust tests
library(phia)
```


## Loading and tidying data
```{r}
project_data <- read_excel("raw_data.xlsx", sheet = 1)

project_data_tidied <- project_data %>% 
  dplyr::select(-X__1, -X__2) %>% 
  replace_with_na_at(.vars = c("CARE1", "CARE2", "CARE3", "CARE4", "CARE5", "CARE6", "CARE7", "CARE8", "CARE9", "CARE10"),
                     condition = ~.x == 6)
```

## Converting data to factors, composites
```{r}
# convert binary data to 2-level factors
project_data_tidied$factored_sex<-factor(project_data_tidied$sex, levels = c(0,1), labels = c("Male", "Female"))

project_data_tidied$factored_condemp<-factor(project_data_tidied$condemp, levels = c(0,1), labels = c("Non-Empathetic", "Empathetic"))

project_data_tidied$factored_condcoat<-factor(project_data_tidied$condcoat, levels = c(0,1), labels = c("No Coat", "Coat"))

# creating mood related composite metric based on row-wise means of PANAS positive and negative values using one of the technique shown here: https://stackoverflow.com/questions/9490485/how-can-i-get-the-average-mean-of-selected-columns
# Composite measure (mean) of the positive mood subscale

project_data_tidied$positive_mood <- rowMeans(subset(project_data_tidied, select = c(PANASPOS1, PANASPOS2, PANASPOS3, PANASPOS4, PANASPOS5, PANASPOS6, PANASPOS7, PANASPOS8, PANASPOS9, PANASPOS10)), na.rm = TRUE)

#project_data_tidied$positive_mood
# Composite measure (mean) of the negative mood subscale
project_data_tidied$negative_mood <- rowMeans(subset(project_data_tidied, select = c(PANASNEG1, PANASNEG2, PANASNEG3, PANASNEG4, PANASNEG5, PANASNEG6, PANASNEG7, PANASNEG8, PANASNEG9, PANASNEG10)), na.rm = TRUE)

#project_data_tidied$negative_mood

project_data_tidied$empathy <- rowMeans(subset(project_data_tidied, select = c(CARE1, CARE2, CARE3, CARE4, CARE5, CARE6, CARE7, CARE8, CARE9, CARE10)), na.rm = TRUE)

project_data_tidied$warmth <- rowMeans(subset(project_data_tidied, select = c(WARM1, WARM2, WARM3, WARM4)), na.rm = TRUE)

project_data_tidied$competence <- rowMeans(subset(project_data_tidied, select = c(COMP1, COMP2, COMP3, COMP4, COMP5)), na.rm = TRUE)
```

## Replicating adjusted means in paper

#### Empathy
Three-way analysis, where `positive_mood` is a covariate:

```{r}
lin_model_empathy_full <- lm(empathy ~ factored_condemp*factored_sex + positive_mood, data = project_data_tidied)

Anova(lin_model_empathy_full, type = 2, adjustment = "bonferroni")
```
Didn't end up using this for calculating the means.

**"Adjusting for mood, we found a significant main effect of nonverbal behavior on ratings of empathy such that participants rated physicians displaying empathic nonverbal behavior as more empathic (M = 3.29, SD = 1.15) than physicians displaying unempathic nonverbal behavior (M = 1.86, SD = .93, F(1,1362) = 568.49, p < .001, η2_p = .30; see Fig 3)."**


```{r}
lin_model_empathy_condemp <- lm(empathy ~ factored_condemp, data = project_data_tidied)

test <- Anova(lin_model_empathy_condemp, type = 2, adjustment = "bonferroni")
test
```

Here, we get $p < .001$ (indicated by the triple star), so that value is verified. Our f-statistic is off--we get 644.55 instead of 568.49.

```{r}
interactionMeans(lin_model_empathy_condemp, adjustment = "bonferroni")
```

We are getting close to their mean for each of these groups. Here, we have the standard error and not the standard deviation. We were not able to replicate their standard deviation, even with manipulation of this standard error.

```{r}
project_data_tidied %>% 
  count(factored_condemp)
```

By hand calculation of standard deviation:
```{r}
0.03960871*sqrt(686) #non-empathetic
0.03941720*sqrt(691) #empathetic
```

---

**"There was also a significant, albeit very small, main effect of subject gender such that male participants rated physicians in both conditions as more empathic (M = 2.70, SD = 1.21) than the female participants (M = 2.49, SD = 1.30, F(1,1362) = 6.60, p = .01, η2_p = .005)."**

```{r}
lin_model_empathy_gender <- lm(empathy ~ factored_sex, data = project_data_tidied)
test_empathy_gender <- Anova(lin_model_empathy_gender, type = 2, adjustment = "bonferroni")
test_empathy_gender
```

Here, we get $p = .002$, which is "double-star", significant, $p<.01$. This matches the paper. Our f-statistic is again off, at 9.5719 instead of 6.60.

```{r}
interactionMeans(lin_model_empathy_gender, adjustment = "bonferroni")
```

Here, we are at the exact adjusted means from the paper for male and female. Again, we have the standard error here instead of standard deviation. We were not able to replicate their standard deviation, even with manipulation of this standard error.

```{r}
project_data_tidied %>% 
  count(factored_sex)
```

By hand calculation of standard deviation:
```{r}
0.05445462*sqrt(541) #males
0.04410946*sqrt(823) #females
```


---

**"There was a marginally significant interaction of nonverbal behavior with participant gender (F(1,1362) = 3.00, p = .084, η2 p = .002) such that in the unempathic condition, women perceived physicians as less empathic (M = 1.73, SD = .88) than men (M = 2.05, SD = .98, F(679) = 13.48, p < .001, η2 p = .02)."**

```{r}
lin_model_condemp_gender_interaction <- lm(empathy ~ factored_condemp*factored_sex, data = project_data_tidied)
test_condemp_gender_interaction <- Anova(lin_model_condemp_gender_interaction, type = 2, adjustment = "bonferroni")
test_condemp_gender_interaction
```


```{r}
phia::interactionMeans(lin_model_condemp_gender_interaction, adjustment = "bonferroni")
```


```{r}
project_data_tidied %>% 
  count(factored_sex, factored_condemp)
```


---

#### Warmth
Three- way analysis, where `positive_mood` is a covariate:
```{r}
lin_model_warmth_full <- lm(warmth~ factored_condemp*factored_sex + positive_mood, data = project_data_tidied)
Anova(lin_model_warmth_full, type = 2, adjustment = "bonferroni")
```
Didn't end using this to calculate the means.

"Adjusting for mood, we found a significant main effect of nonverbal behavior on ratings of warmth such that participants rated physicians displaying empathic nonverbal behavior as more warm (M = 3.73, SD = .93) than physicians displaying unempathic nonverbal behavior (M = 2.28, SD = .98, F(1,1362) = 674.49, p < .001, η2 p = .33)."

```{r}
lin_model_warmth_condemp <- lm(warmth ~ factored_condemp + positive_mood, data = project_data_tidied)
Anova(lin_model_warmth_condemp, type = 2, adjustment = "bonferroni")

interactionMeans(lin_model_warmth_condemp, adjustment = "bonferroni")
```

"We also found a significant interaction of nonverbal behavior with participant gender (F(1,1362) = 4.88, p = .027, η2 p = .004) such that in the unempathic condition, women perceived physicians as less warm (M = 2.20, SD = 0.98) than men (M = 2.42, SD = .95, F(1,679) = 4.46, p = .035, η2 p = .007)."

```{r}
lin_model_warmth_condemp_gender_interaction <- lm(warmth ~ factored_sex*factored_condemp, data = project_data_tidied)
Anova(lin_model_warmth_condemp_gender_interaction, type = 2)

interactionMeans(lin_model_warmth_condemp_gender_interaction)
```


```{r}
0.03633678*sqrt(686)
0.03621157*sqrt(691)
```

#### Competence
Three-way analysis, where `positive_mood` is a covariate:
```{r}
lin_model_competence_full <- lm(competence ~ factored_condemp*factored_sex + positive_mood, data = project_data_tidied)

Anova(lin_model_competence_full, type = 2, adjustment = "bonferroni")
```
Didn't end up using this for calculating the means

**"Adjusting for mood, there was a significant main effect of nonverbal behavior on ratings of competence, such that participants rated physicians displaying empathic nonverbal behavior as more competent (M = 3.64, SD = .65) than physicians displaying unempathic nonverbal behavior (M = 3.21, SD = .81, F(1,1362) = 85.11, p < .001, η2_p = .06)."**

```{r}
lin_model_competence_condemp <- lm(competence ~ factored_condemp + positive_mood, data = project_data_tidied)
Anova(lin_model_competence_condemp, type = 2, adjustment = "bonferroni")
```

```{r}
interactionMeans(lin_model_competence_condemp, adjustment = "bonferroni")
```

**"Positive and negative mood were also significant covariates of these effects, such that higher positive mood was associated with higher ratings of physician competence (F(1,1362) = 87.65, p < .001, η2_p = .06) and higher negative mood was associated with lower ratings of physician competence (F(1,1362) = 14.25, p < .001, η2_p = .01)."**

```{r}
lin_model_competence_pos_mood <- lm(competence ~ positive_mood, data = project_data_tidied)
Anova(lin_model_competence_pos_mood, type = 2, adjustment = "bonferroni")
```

```{r}
lin_model_competence_neg_mood <- lm(competence ~ negative_mood, data = project_data_tidied)
Anova(lin_model_competence_neg_mood, type = 2, adjustment = "bonferroni")
```

## Compiled means and errors for replicating fig 3

```{r}
emp_mean <- as.data.frame(interactionMeans(lin_model_empathy_condemp, adjustment = "bonferroni"))[1,2]
emp_error <- as.data.frame(interactionMeans(lin_model_empathy_condemp, adjustment = "bonferroni"))[2,2] 
emp_mean
emp_error

warmth_mean<- as.data.frame(interactionMeans(lin_model_warmth_condemp, adjustment = "bonferroni"))[1,2]
warmth_error<- as.data.frame(interactionMeans(lin_model_warmth_condemp, adjustment = "bonferroni"))[2,2]
warmth_mean
warmth_error

competence_mean<-as.data.frame(interactionMeans(lin_model_competence_condemp, adjustment = "bonferroni"))[1,2]
competence_error<-as.data.frame(interactionMeans(lin_model_competence_condemp, adjustment = "bonferroni"))[2,2]
competence_mean
competence_error

#all adjusted for mean.
#error bars * 1.96, -1.96 for confidence intervals
```

http://www.cookbook-r.com/Graphs/Plotting_means_and_error_bars_(ggplot2)/



## Bar Plot in Paper
```{r}

# build comparison dataframe
emp_graph <- interactionMeans(lin_model_empathy_condemp, adjustment = "bonferroni") %>%
  mutate(DV = "Empathy")
wrm_graph <- interactionMeans(lin_model_warmth_condemp, adjustment = "bonferroni") %>%
  mutate(DV = "Warmth")
cmp_graph <- interactionMeans(lin_model_competence_condemp, adjustment = "bonferroni") %>%
  mutate(DV = "Competence")
grr <- rbind(emp_graph,wrm_graph,cmp_graph)


# Ensure graph components are matching color and position of papers
grr$DV <- factor(grr$DV,levels = c("Empathy", "Warmth", "Competence"))

# plot parameterization
intrabar_spacing = -.9 # how far red and blue bars are from eachother
ggplot(grr, aes(x=DV, y=`adjusted mean`, fill=factored_condemp, ylim(1,5), order = -as.numeric(factored_condemp))) + 
    geom_bar(position=position_dodge(width = intrabar_spacing), stat="identity", width = .8) +
    # Use 95% confidence intervals
    geom_errorbar(aes(ymin=`adjusted mean`- `std. error`*1.96, ymax=`adjusted mean`+ `std. error`*1.96),
                  width=.075,  # Width of the error bars
                  position=position_dodge(width =intrabar_spacing)) + 
    coord_cartesian(ylim=c(1,5))+
  scale_fill_manual(values=c("blue","red")) # match paper color
  
```

## Attempted ANCOVA plots
```{r}

mod <- ancova( data=project_data_tidied, empathy ~ factored_condemp*factored_sex)
pred <- predict(mod)

```



























